The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-02-17 paper.tex
TPs and AIAs examples

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
%\newcommand{\preview}{\textnormal{\ident{preview}}}
%\newcommand{\links}{\textnormal{\ident{links}}}
%\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}

@@

\vspace{-0.25em}
\begin{itemize}
  \item $\choices(\sigma) = (\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n)$ is
    an operation that takes a state and generates options designated by an identifier $\iota_i$
    and represented by a state $\sigma_i$,
  \item $\select(\sigma) = e$ is an operation that returns generated program for a given state.
\end{itemize}
\end{definition}

@@

\subparagraph{Formal model.}
The options generated by The Gamma let the user select both the next operation and the parameters
of the previously selected operation. The available operations and parameters are generated
based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
contains the current schema $S$ and the operations applied so far. In the following, we write
$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
\[
\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
\]
The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
expects further parameters or whether it is fully-specified. In the first case, the recommendation
engine generates possible additional parameter values $p_', p_'', \ldots$ based on the schema $S$,
the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
The $\choices$ operation then generates options that add the additional parameter. We generated the
identifiers $\iota_',\iota_'',\ldots$ based on the state and the parameter value, such as
\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
\[
\begin{array}{l}
\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots
\end{array}
\]
If the last operation takes no further parameters, the system produces a choice
of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
operation then returns options that add the additional operation:
\[
\begin{array}{l}
\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots
\end{array}
\]
Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
host language lets us compose functions using $f\circ g$, we can write:
\[
\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
\]
The recommendation engine behind The Gamma provides a domain-specific logic for generating
possible operations and their parameters based on the current schema of the data. As the above
definition shows, this underlying engine can be easily exposed through the common
choose-your-own-adventure interface.

\subsection{Type providers}

revisit

\subsection{AI assistants}

@@

>>>>

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
%\newcommand{\preview}{\textnormal{\ident{preview}}}
%\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\vspace{-0.25em}
\raggedright
\begin{itemize}
  \item $\choices(\sigma) = \{\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n\}$ is
    an operation that takes a state and \\ generates options designated by an identifier $\iota_i$
    and represented by a state $\sigma_i$,
  \item $\select(\sigma) = e$ is an operation that returns generated program for a given state.
\end{itemize}
\end{definition}

@@

\subparagraph{Formal model.}
The options generated by The Gamma let the user select both the next operation and the parameters
of the previously selected operation. The available operations and parameters are generated
based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
contains the current schema $S$ and the operations applied so far. In the following, we write
$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
\[
\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
\]
The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
expects further parameters or whether it is fully-specified. In the first case, the recommendation
engine generates possible additional parameter values $p_', p_'', \ldots$ based on the schema $S$,
the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
The $\choices$ operation then generates options that add the additional parameter. We generated the
identifiers $\iota_',\iota_'',\ldots$ based on the state and the parameter value, such as
\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
\[
\begin{array}{l}
\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
\qquad\;\;    \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots~\}
\end{array}
\]
If the last operation takes no further parameters, the system produces a choice
of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
operation then returns options that add the additional operation:
\[
\begin{array}{l}
\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
\qquad \;\; \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots~\}
\end{array}
\]
Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
host language lets us compose functions using $f\circ g$, we can write:
\[
\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
\]
The recommendation engine behind The Gamma provides a domain-specific logic for generating
possible operations and their parameters based on the current schema of the data. As the above
definition shows, this underlying engine can be easily exposed through the common
choose-your-own-adventure interface.

\subsection{Type providers}
The type provider mechanism in F\# operates at the level of the type system. It is not a merely an
editor feature. A type provider for a data source, such as the World Development Indicators database,
generates a collection of types that model the external data source. In F\#, the types are classes
with members that implement the logic to retrieve data at runtime.

The auto-completion mechanism in F\# code editors, which implements the choose-your-own-adventure
interaction pattern, is not specific to type providers. It offers a list of members of an object
based on its type. We model the completion as an iterative process, repeatedly adding further
members to a chain. The state $\sigma$ thus consists of an initial expression on which the
completion is invoked, a chain of selected members and the type of the last member.

To model the completion mechanism, we also need to model information about types. We loosely
follow the Foo calculus model \cite{petricek-2016-fsdata} and write $\mathbb{C}$ for a set
of class definitions, each consisting of an implicit class constructor and a collection of
members $M$:
\[
\begin{array}{rcl}
\sigma &=& e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C\\
\mathbb{C} &=& \{ C \mapsto \ident{type}~C(\overline{x:\tau}) = \overline{M},~\ldots~ \}\\
M &=& \ident{member}~\iota\!:\!C=e
\end{array}
\]
Each member in the Foo calculus consists of a of a name $\iota$, return type $C$ and implementation
$e$. For our purposes, we only need the type information and so the operations that define the
choose-your-own-adventure are parameterized by the set of classes $\mathbb{C}$.

The $\choices_\mathbb{C}$ operation finds the class definition corresponding to the type of the
last member in the current chain. It offers choices appending each of the available
members to the current chain. The $\select_\mathbb{C}$ operation returns the constructed
member chain:

\[
\begin{array}{l}
\choices_\mathbb{C}(e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C) =\\
\qquad \{\; \iota' \mapsto (e . \iota_1 . \lbrack\ldots\rbrack . \iota_n . \iota', C')\\
\qquad \;\; \iota'' \mapsto (e . \iota_1 . \lbrack\ldots\rbrack . \iota_n . \iota'', C''), ~\ldots~\}\\[1em]
\end{array}
\]\vspace{-2em}
\[
\begin{array}{rcl}
\quad \textnormal{where}~\mathbb{C}(C) &=& \ident{type}~C(\overline{x:\tau}) = M', M'', \ldots\\
\quad \textnormal{and}~M' &=& \ident{member}~\iota'\!:\!C'=e'
\end{array}
\]\vspace{-0.5em}
\[
\select_\mathbb{C}(e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C) = e . \iota_1 . \ldots . \iota_n
\]
The model does not directly refer to type providers. Those are responsible solely for generating
the type definitions in $\mathbb{C}$ as documented in earlier work \cite{petricek-2016-fsdata}.
It is worth noting that the type provider for data exploration, implemented by The Gamma,
additionally needs to generate classes lazily \cite{petricek-2017-dotdriven}. To model this aspect,
the simple lookup $\mathbb{C}(C)$ needs to be replaced with an operation that returns the type
definition, alongside with a new context $\mathbb{C}'$ that contains additional generated
type definitions (return types for all the members of the class~$C$).

The model follows the internal mode of embedding the interaction in the program. It is easy to
define the \ident{decode} operation that takes the resulting generated expression and returns the
sequence of choices, because the choices are items of the member chain. A slight caveat is that
the completion is not invoked on an empty hole, but on a hole that contains the initial expression
on which the completion is applied. We can model this using filled holes \cite{omar-2019-holes}
and write $?_e$ for a hole containing the initial expression $e$. The $\ident{init}(?_e)$ operation
then returns $e$ alongside with an empty chain and the type of $e$.

As noted earlier, The Gamma does not embed query expressions directly into the generated code.
It uses the same model as type providers and generates choices as members of
types behind the scenes. We return to the differences between the two models in
Section~\ref{sec:properties}.

\subsection{AI assistants}
AI assistants guide the analyst through a data wrangling task. They
generate a data cleaning script, taking into account constraints selected by the user.
Most AI assistants obtain the script by performing statistical optimization with respect
to a set of constraints specified by the user. That is, they look for an expression from the
set of all possible expressions that optimizes some objective function that assigns score to
the expression with respect to the given input data. Note that AI assistants do not iterate
over all possible expressions. They use a machine learning method to approximate a solution
to the problem.

An optimization-based AI assistant \cite{petricek-2023-aias} thus provides another, very
different, way of implementing the choose-your-own-adventure pattern. The assistant operates
with respect to some input data $X$ that does not change during the interaction and so we
parameterize the choose-your-own-adventure calculus operations by the data.
The input data $X$ can be actual input data or a representative sample and so the AI assistant
can be use past data to infer a cleaning script that will be used on new inputs.

The state $\sigma$ consists of a set of constraints specified by the user. We write
$c$ for individual constraints and $\boldsymbol{c}$ for a set of constraints. The initial
state is an empty set:
\[
\begin{array}{rcl}
\sigma &=& \{ c_1, \ldots, c_n \}\\
\sigmaN &=& \emptyset
\end{array}
\]
Unlike in the previous examples, the crucial logic of an AI assistants is implemented in the
$\select$ operation. The operation runs the optimization algorithm to choose the best cleaning
script for given constraints. Formally, this can be written using the $\argmax$ operator which
finds an argument (an expression) for which the given function (scoring function) is maximized.
The user-specified constraints can either restrict the set of possible expressions or influence
the scoring function. More formally, we assume that:

\begin{itemize}
\item $E_{\boldsymbol{c}}\subseteq E$ is a set of expressions that respect constraints $\boldsymbol{c}$,
\item $Q_{\boldsymbol{c}}(X, e)$ is a scoring function with respect to the constraints $\boldsymbol{c}$,
  which returns the score of an expression $e$, i.e., how good $e$ is at cleaning the data $X$.
\end{itemize}

\noindent
For a given set of constraints $\boldsymbol{c}$, the $\select$ operation looks
for $e\in E_{\boldsymbol{c}}$ with the largest score:
\[
\select_X(\boldsymbol{c}) = \argmax_{e \in E_{\boldsymbol{c}}} Q_{\boldsymbol{c}}(X, e)
\]
The actual implementation of the optimization uses various machine learning techniques to
find the optimal expression. In case of datadiff, $X$ is a pair of datasets $X_1, X_2$ to be
reconciled. The AI assistant uses the Hungarian algorithm \cite{sutton-2018-datadiff} to construct
a matching of columns from $X_1$ and $X_2$. The generated expression is a sequence of patches
that can be applied to $X_2$ in order to reconcile its structure with the sturcture of $X_1$.
The constraints specified by the user restrict the space of possible column matchings and so they
affect $E_\boldsymbol{c}$. The scoring function $Q_\boldsymbol{c}$ is independent of
the constraints and computes a sum of distances between the statistical distributions of the
columns from $X_1$ and a patched version of $X_2$.

The $\choices$ operation is responsible for generating possible constraints that the user may
want to add to guide the inference. AI assistants typically offer the user options to prevent
or adapt some aspect of the cleaning logic inferred by the system. For example, if datadiff
matches two columns, it will offer a constraint to prevent the matching. It also generates
constraints that let the user force a specific matching.

To implement $\choices_X$, optimization-based AI assistants first call $\select_X(\sigma)$ to
get the best expression $e$. Based on this, they generate possible constraints $c_1, c_2, \ldots$
that the user may want to choose from. The identifiers $\iota_1,\iota_2, \ldots$ provide a
human-readable description of the constraints. Note that this operation is specific to the particular
AI assistant. The $\choices_X$ operation then offers a list of constraint sets where the additional
constraint is added to the previously collected set:
\[
\choices_X(\boldsymbol{c}) = \{\iota_1 \mapsto \boldsymbol{c}\cup\{ c_1 \}, \iota_2\mapsto\boldsymbol{c}\cup\{ c_2 \}, \ldots\}
\]
The integration of an AI assistant, as described here, has to follow the external mode of embedding.
The interaction with the assistant results in a cleaning script (expression), but there is no way
of reconstructing the constrains used to guide the optimization. To support internal embedding, the
$\select$ operation would need to explicitly include the constraints in the resulting expression.
However, rerunning the $\select$ operation with the same constraints may result in a different
cleaning script if the machine learning algorithm is probabilistic.

@@

\section{Limitations}

hard to edit - you cannot easily go back
